---
title: "Econometrics 🫀"
subtitle: "<span class=\"subtitle-text\">TA Session 2</span>"
author: "Lucia Sauer"
institute: "<em>Barcelona School of Economics</em>"
date: "2025-10-01"
output:
  html_document:
    theme: readable
    highlight: haddock #default, tango, pygments, kate, monochrome, espresso, zenburn, haddock
    df_print: paged
---


In this notebook we will review:

1. Conditional means
2. OLS in matrix algebra
3. OLS using R and Python preset functions
4. Plotting observations and fitted line
5. Numerical properties of OLS estimators



```{r}
#install libraries
# install.packages("wooldridge")
# install.packages("DT")
# install.packages("dplyr")
#install.packages("dplyr")
# install.packages("ggplot2")
```

---

# **1. Conditional Means 🏡**
Using the `hprice1` dataset from the `wooldridge` package, we will explore the conditional mean of house prices given the number of bedrooms.

-  `price`     House price ($1000s)         
-  `assess`    Assessed value ($1000s)      
-  `bdrms`     Number of bedrooms           
-  `lotsize`   Lot size (square feet)       
-  `sqrft`     House size (square feet)     
-  `colonial`  =1 if home is colonial style 
-  `lprice`    Log(price)                   
-  `lassess`   Log(assess)                  
-  `llotsize`  Log(lotsize)                 
-  `lsqrft`    Log(sqrft)                   


```{r}
library(wooldridge)
#load dataset

```


Calculate mean of house prices conditional on number of bedrooms.

$$\hat{E} \ [price|bdrms]$$

```{r,message=FALSE}
library(dplyr)

```

## Plot the conditional mean

```{r,message=FALSE, warning=FALSE}
library(ggplot2)
#initialize the plot

  # Scatter of raw data
  
  
  # Line plot of conditional means
  
  
  # Titles and labels
  
  
  # Styling similar to your Python example
  # theme_minimal(base_size = 14) +
  # theme(
  #   panel.grid.minor = element_blank(),
  #   panel.grid.major = element_line(color = "grey80"),
  #   axis.line = element_line(color = "black"),
  #   axis.ticks = element_line(color = "black"),
  #   plot.title = element_text(size = 16, face = "bold")
  # )
```

---

# 2. OLS in Matrix algebra 🧮

Using the same dataset as before, we will estimate the following model:

$$price\_i = \beta_1 + \beta_2 \cdot bdrms_i  + \beta_3 \cdot sqrft_i + \beta_4 \cdot colonial_i + \epsilon_i $$

We can write this model in matrix form as:
$$ Y = X \beta + \epsilon $$

Where:

- $Y$ is a $n \times 1$ vector of the dependent variable (house price)
- $X$ is a $n \times k$ matrix of the independent variables (including a column of ones for the intercept)
- $\beta$ is a $k \times 1$ vector of coefficients



## 2.1 Estimate OLS coefficients using matrix algebra
```{r,message=FALSE}
#define X and y


#define X_tX and X_ty


#solve the system

```

Fitted model:

$$\hat{price}_i = -21.55 + 12.49\cdot bdrms_i  + 0.13\cdot sqrft_i + 13.08\cdot colonial_i$$

where the dependent variable `price` is in **$1000s**.

## 2.2 Calculate fitted values, SSE, and $R^2$

$$\hat{y} = X \hat{\beta}$$

```{r,message=FALSE}
#Calculate y_hat

```

$$SSE = \sum_{i}^{n} (y_i - \hat{y}_i)^2 = \sum_{i}^{n} \hat{\epsilon}^2 = \hat{\epsilon}´\hat{\epsilon}$$

```{r,message=FALSE}
#epsilon_hat, SSE


```

$$R^2 = 1 - \frac{SSE}{SST}$$

where 
$$SST = \sum_{i}^{n} (y_i - \bar{y}_i)^2 = (y - \mathbf{1} \bar{y})'(y - \mathbf{1} \bar{y})$$

```{r,message=FALSE}
#SST, R^2


```

Note: 

- The SSE is expressed in squared thousands of dollars.
- The R² is unitfree, and tells us that about 64% of the variation in house prices is explained by the model.

---

# **3. OLS using R and Python preset functions**

We can also estimate the same model using preset functions in Python and R.

- Python: `statsmodels` package
- R: `lm()` function

```{r,message=FALSE}
#Estimate the model with lm function 


```

---

# **4. Plotting observations and fitted line**

For a simple model of $K=2$, estimate the model and plot the observations and the fitted line.

$$ price_i = \beta_1 + \beta_2 \cdot sqrft_i + \epsilon_i $$
```{r,message=FALSE}
#first we need to estimate simple model

```


```{r,message=FALSE}
#plot observations with scatter and fitted line

    #scatter with raw data
    
    #fitted line
    
    #labels and styling
    
  #   theme_minimal(base_size = 14) +
  #   theme(
  #   panel.grid.minor = element_blank(),
  #   panel.grid.major = element_line(color = "grey80"),
  #   axis.line = element_line(color = "black"),
  #   axis.ticks = element_line(color = "black"),
  #   plot.title = element_text(size = 16, face = "bold")
  # )
```

---

# **5. Numerical properties of OLS estimators:**


1. The sum of the residuals is zero:
$$\sum_i ^ n\hat{\epsilon}_i = \sum_i ^ n (y_i - \hat{y}_i)=0$$

on average, the residuals cancel out.

```{r,message=FALSE}
#check sum of epsilon_hat = 0

```

```{r,message=FALSE}
# Create a data frame with index and residuals

#inspect visually this property

  # theme_minimal(base_size = 14) +
  # theme(
  #   panel.grid.minor = element_blank(),
  #   panel.grid.major = element_line(color = "grey80"),
  #   axis.line = element_line(color = "black"),
  #   axis.ticks = element_line(color = "black"),
  #   plot.title = element_text(size = 16, face = "bold")
  # )
```

2. OLS unit dependent, hence SSE also unit dependent

Our dependent variable is expressed in \$1000s, so let's scale it to actual dollars and see how the SSE changes.


$$1000*price = \beta X + \epsilon$$

$$\hat{\beta} ^{*} = (X'X)^{-1}X'price*1000$$
$$\hat{\beta} ^{*} = 1000\hat{\beta}$$


```{r,message=FALSE}
# df$price <- df$price * 1000
# model_dollars <- lm(price ~ sqrft, data = df)
# beta_hat_dollars <- coef(model_dollars)
# sse_dollars <- sum(residuals(model_dollars)^2)
# cat("Coefficients in thousands of dollars:", round(beta_hat, 4), "\n")
# cat("Coefficients in dollars:", round(beta_hat_dollars, 4), "\n")

# cat("SSE in thousands of dollars:", round(sse, 2), "\n")
# cat("SSE in dollars:", round(sse_dollars, 2), "\n")

```